\section{Homogeneous polynomials in 2 variables}\label{sec:part2}
In the preceding section, we considered degree $2$ homogeneous polynomials in $n$ variables. Now we will instead consider a situation involving degree $n$ homogeneous polynomials in $2$ variables. Although we will set things up quite differently, the driving idea is once again Lemma \ref{lem:FT-conical-subset} which reduces our task to analyzing hyperplane cross-sections. The cross-sections in this setting will have a particularly natural description, revealed in \eqref{eq:part2-cross-section-soln-count}.

First, let us define the set $S$ of interest. Let $\F_p[x,y]_n$ denote the space of homogeneous polynomials of degree $n$ over $\F_p$. It is an $(n+1)$-dimensional vector space, spanned by the basis elements
\[
	\{x^n, x^{n-1}y, \ldots, xy^{n-1}, y^n\}.
\]
Define $V$ to be the dual space of this vector space:
\[
	V \coloneqq \left(\F_p[x,y]_n\right)^*.
\]
Of particular algebraic interest are the elements of $V$ which are given by \emph{evaluation} at a point---these are the functionals $\varepsilon_{a,b}\colon \F_p[x,y]_n \to \F_p$ of the form
\[
	\varepsilon_{a,b}(P) = P(a,b) 
\]
for some $a,b\in \F_p$.

However, the set of all $\varepsilon_{a,b}$ need not be conical: consider
\[
	2\varepsilon_{1,0} \in \F_3[x,y]_2.
\]
This functional at $P(x,y) = x^2$ takes the value $2$, but $2\in \F_3$ is not a square and thus $2\varepsilon_{1,0}$ cannot be equal to $\varepsilon_{a,b}$ for any $a,b\in \F_3$. To get around this, we will instead let $S$ be the set
\begin{equation}\label{eq:part2-defn-S}
	S=\left\{c\varepsilon_{a,b} \mid a,b,c \in \F_p\right\}.
\end{equation}
The Fourier transform $\widehat{\IND_S}$ will be a $\C$-valued function on $V^*$, where $V^*$ is the double dual of $\F_p[x,y]_n$. Since $\F_p[x,y]_n$ is finite-dimensional, it is canonically isomorphic to its double dual, so we will identify $V^* = \F_p[x,y]_n$.

Define $\P^1$ to be the set of lines through the origin in $\F_p^2$. For each line $\ell \in \P^1$, pick a nonzero point $(a_\ell,b_\ell)$ on $\ell$. Then we have
\begin{equation}\label{eq:part2S}
	S = \{0\} \cup \left\{c\varepsilon_{a_\ell,b_\ell} \mid c\in \F_p \text{ nonzero}, \ell \in \P^1 \right\}
\end{equation}
and hence
\[
	|S| = 1 + (p+1)(p-1) = p^2.
\]
We will now give a very simple description of $\widehat{\IND_S}$.
\begin{thm}\label{thm:part2thm}
	Let $\F_p[x,y]_n$ be the space of homogeneous polynomials of degree $n$ over $\F_p$, $V$ its dual space, and $S\subset V$ the set defined by \eqref{eq:part2-defn-S}.
	
	An element $\xi \in V^*$ is a polynomial $P(x,y)$. Let $N_\xi$ be the number of solutions to $P(x,y)=0$ in $\F_p^2$. Then
	\begin{equation}
		\widehat{\IND_S}(\xi) = \frac{pN_\xi - p^2}{p-1}.\label{eq:FTpart2}
	\end{equation}
\end{thm}
\begin{proof}
	An element $c\varepsilon_{a,b}\in S$ is in $\ker \xi$ when $P(a,b) = 0$. From \eqref{eq:part2S}, we have
	\begin{equation}\label{eq:part2-cross-section-soln-count}
		|S \cap \ker \xi| = N_\xi
	\end{equation}
	and thus, by Lemma \ref{lem:FT-conical-subset},
	\begin{align*}
		\widehat{\IND_S}(\xi) &= \frac{p|S\cap \ker \xi| - |S|}{p-1}\\
		&= \frac{pN_\xi - p^2}{p-1}.\qedhere
	\end{align*}
\end{proof}
There is unfortunately no explicit formula for the number $N_\xi$ of solutions to $P(x,y) = 0$ in terms of the coefficients of $P$, so we will not attempt to expand further on \eqref{eq:FTpart2}. Instead, we offer an alternative geometric perspective which also leads to a proof of Theorem \ref{thm:part2thm}. Each polynomial in $V^* = \F_p[x,y]_n$ can be thought of a row vector $(a_0,\ldots,a_n)$ where $a_i$ is the coefficient of $x^{n-i}y^i$ in the polynomial. Then ``evaluation at $(a,b)$'' is given by pairing with the column vector
\[
	(a^n,a^{n-1}b,\ldots,ab^{n-1},b^n)^\top.
\]
With this identification, $S = \{\xi_{a,b,c}\}$ is the union of the following lines:
\begin{align*}
	\ell_\infty &= \langle (0,0,\ldots,0,1)^\top \rangle,\\
	\ell_0 &= \langle (1,0,\ldots,0,0)^\top \rangle,\\
	\ell_1 &= \langle (1,1,\ldots,1,1)^\top \rangle,\\
	\ell_2 &= \langle (1,2,\ldots,2^{n-1},2^n)^\top \rangle,\\
	\vdots &= \vdots\\
	\ell_{p-1} &= \langle (1,p-1,\ldots,(p-1)^{n-1},(p-1)^n)^\top \rangle
\end{align*}
As an aside for the reader familiar with algebraic geometry, if we think of $S$ as living inside projective space $\P^{n}$, it is the \emph{rational normal curve} of degree $n$ over $\F_p$.

Using the fact that $S$ is the union of the above lines, we can write
\begin{align*}
	\widehat{\IND_S} &= \sum \widehat{\IND_{\ell_i}} - p\widehat{\IND_{\{0\}}} \\
	&= p\left(\sum \IND_{\ell_i^\bot} - \IND_{V^*}\right).
\end{align*}
where we have used the linearity of the Fourier transform together with Theorem \ref{thm:FT-subspace}. A polynomial $P$ is in $\ell_i^\bot$ when it vanishes on $\ell_i$. If $P$ vanishes at $N$ points in total, then it vanishes on $(N - 1)/(p-1)$ of the lines $\ell_i$. This gives another proof of Theorem \ref{thm:part2thm}.

What is the ``geometry'' of these hyperplanes $\ell_i^\bot$? If $n < p$ then clearly not all $p+1$ of the lines $\ell_i$ can be linearly independent. We claim, however, that every subset of size at most $n+1$ is indeed linearly independent. A way of seeing this is by noting that the Vandermonde determinant is nonzero. Alternatively, one can explicitly produce a polynomial $P \in \F_p[x,y]_n$ which vanishes on $n$ given lines but not on the others. On the other hand, if an element of $\F_p[x,y]_n$ vanishes on $n+1$ of the lines, then in fact it vanishes on all of the lines. This is entirely what we expect: dividing the polynomial by $y^n$ gives a univariate polynomial in $(x/y)$ of degree at most $n$. If such a polynomial has $n+1$ roots, then it is identically zero.

In this section, we considered a conical set $S$ arising from evaluation functionals. The Fourier transform $\widehat{\IND_S}$ records the number $N_\xi$ of solutions to polynomials $\xi\in\F_p[x,y]_n$ in the manner given by Theorem \ref{thm:part2thm}. As we just discussed above, the only possible values of $N_\xi$ are $(p-1)k + 1$, where $k$ denotes the number of lines $\ell_i$ on which the polynomial $\xi$ vanishes. In particular, $k\in\{0,\ldots,n\}$ for nonzero $\xi$ and $k=p+1$ for $\xi = 0$.